Counting And Confusion: Bayesian Rate Estimation With Multiple Populations

### 2013-01-16 many.tex
Edited III.B and IV.A for clarity.

For the usual $k=1$ case \citep{Biswas2009}, the marginalized
posterior (Eq.~\eqref{eq:rate-shape-posterior}) becomes
\begin{multline}\label{LEeq1}
p_{\rm LE}\left( R_f, R_b, \theta | d \right) \propto \left(R_f
\hat{f}(x_N,\theta) + R_b \hat{b}(x_N,\theta)\right) \\ \times \exp\left[-\left(
  R_f (1-\hat{F}(x_N,\theta)) + R_b (1-\hat{B}(x_N,\theta)) \right)
  \right] \\ \times \frac{p(\theta)}{\sqrt{R_f R_b}}.
\end{multline}
where $x_N$ denotes the loudness of the loudest event, and $R_f$ and
$R_b$ are the number of events expected \emph{above our original
  threshold} (so, for example, $R_f (1 - \hat{F}(x_N,\theta))$ is the
number of foreground events expected above loudness $x_N$).  In the
loudest event statistic paper~\citep{Biswas2009}, the authors assume
the background distribution and rate are known, which corresponds to
using a delta function prior on $R_b$. They further assume a flat
prior on $R_f$ and that the foreground and background distributions do
not depend on any unknown free parameters. With these assumptions, the
posterior on $R_f$, Eq.~\eqref{LEeq1}, is modified to
\begin{multline}
p_{\rm LE}(R_f|d) \propto  \left(R_f
\hat{f}(x_N) + R_b \hat{b}(x_N)\right) \\ \times \exp\left[-\left(
  R_f (1-\hat{F}(x_N)) + R_b (1-\hat{B}(x_N)) \right)
  \right] .
\end{multline}
Integrating over $R_f$ gives
\begin{multline}
\int_0^\infty p_{\rm LE}(R_f|d) {\rm d}R_f =  \frac{R_b \hat{b}(x_N)}{(1-\hat{F}(x_N))} {\rm e}^{-(1-\hat{B}(x_N)) R_b}\\
%\exp\left[-(1-\hat{B}(x_N))R_b\right] \\ 
\times \left(\frac{\hat{f}(x_N)}{(1-\hat{F}(x_N)) R_b \hat{b}(x_N)} + 1 \right)
\end{multline}
and so the normalised posterior is
\begin{multline}
p_{\rm LE}(R_f|d) = \frac{(1-\hat{F}(x_N))}{1+\Lambda} \left( 1+ R_f (1-\hat{F}(x_N))\Lambda \right) \\
\times \exp\left[-R_f (1-\hat{F}(x_N))\right]
\end{multline}
in which we have defined
\begin{equation}
\Lambda \equiv \frac{\hat{f}(x_N)}{(1-\hat{F}(x_N)) R_b \hat{b}(x_N)}.
\end{equation}
With the further identification $\mu \equiv R_f$ and $\hat{\epsilon} \equiv 1 - \hat{F}(x_N)$, this is Eq.~(18) of~\citep{Biswas2009} and we have shown how their parameter $\Lambda$ is related to the foreground and background distributions used here.

@@

\subsection{Foreground dominated statistic}
If we set the threshold for including an event, $x_{\rm min}$,
sufficiently high, we can ensure that $\hat{f}(x_i,\theta) \gg
\hat{b}(x_i,\theta)$ for all ranking statistics $x_i$ in the data
set. The posterior can then be approximated by
\begin{multline}
p_{\rm FD}\left( R_f, R_b, \theta | d \right) \\ \propto \prod_{i} \left[
  \hat{f}\left(x_i, \theta\right)\right] R_f^N \exp\left[-\left( R_f +
  R_b \right) \right] \frac{p(\theta)}{\sqrt{R_f R_b}}.
\end{multline}
Normalization over $R_b$ gives a constant factor and the posterior on
the foreground rate becomes
\begin{multline}
p_{\rm FD}\left( R_f, \theta | d \right) \\ \propto \prod_{i} \left[
  \hat{f}\left(x_i, \theta\right)\right] R_f^{N-\frac{1}{2}}
\exp\left[-R_f \right] p(\theta).
\end{multline}
Ignoring the dependence on $\theta$, this is peaked at a rate $R_f =
N-1/2$, so we have the expected result that, in the foreground
dominated regime, the rate is approximately equal to the number of
events observed (the $1/2$ comes from our use of the Jeffreys prior on
the rate).  We note that the rate $R_f$ is defined as the rate of
events occurring above the specified threshold and therefore if the
foreground dominated statistic were used to estimate rates $R_{f,1}$
and $R_{f,2}$ for thresholds $x_{\rm min}=x_1$, $x_{\rm min}=x_2$
respectively, these rates should be compared by equating $R_{f,1}
(1-\hat{F}(x_2,\theta))$ and $R_{f,2} (1-\hat{F}(x_1,\theta))$.

@@

To explain this, we will use the same model as described in the
previous subsection, and we will begin with a very specific example.
Let $\Lambda_f = 1$, meaning that the expected number of actual events
with $x > x_1$ is one.  Then there is a $1/64$ chance ($\approx
1.6\%$) that the loudest event will have $x_{LE} > 4 x_1$.  Consider
this case, and let us also assume that there are no events (noise or
actual) with $x_1 < x < x_{LE}$.  The loudest event estimate basically
``throws away'' the information that there are no events in this
interval, which (we show below) leads an estimate
$\Lambda_f^\mathrm{LE} \approx 64$, which is $64$ times the true rate!
Now, if the true rate really were $\Lambda_f = 64$, then the expected
number of events with $x > x_1$ would be $64$.  So in this case, the
loudest event estimate ignores the fact that there are $\sim 56 -72$
``missing'' events.  However a Bayesian estimate with $x_\mathrm{th}$
set to $x_1$ incorporates this information quite naturally, and so
(correctly) yields an estimated $\Lambda_f$ of order one.

>>>>

For the usual $k=1$ case \citep{Biswas2009}, the marginalized
posterior (Eq.~\eqref{eq:rate-shape-posterior}) becomes
\begin{multline}\label{LEeq1}
p_{\rm LE}\left( R_f, R_b, \theta | d \right) \propto \left(R_f
\hat{f}(x_N,\theta) + R_b \hat{b}(x_N,\theta)\right) \\ \times \exp\left[-\left(
  R_f (1-\hat{F}(x_N,\theta)) + R_b (1-\hat{B}(x_N,\theta)) \right)
  \right] \\ \times \frac{p(\theta)}{\sqrt{R_f R_b}}.
\end{multline}
where $x_N$ denotes the loudness of the loudest event, and $R_f$ and
$R_b$ are the number of events expected \emph{above our original
  threshold} (so, for example, $R_f (1 - \hat{F}(x_N,\theta))$ is the
number of foreground events expected above loudness $x_N$).  In the
loudest event statistic paper~\citep{Biswas2009}, the authors assume
the background distribution and rate are known, which corresponds to
using a delta function prior on $R_b$. They further assume a flat
prior on $R_f$ and that the foreground and background distributions do
not depend on any unknown free parameters. With these assumptions, the
posterior on $R_f$, Eq.~\eqref{LEeq1}, is modified to
\begin{multline}
p_{\rm LE}(R_f|d) \propto  \left(R_f
\hat{f}(x_N) + R_b \hat{b}(x_N)\right) \\ \times \exp\left[-\left(
  R_f (1-\hat{F}(x_N)) + R_b (1-\hat{B}(x_N)) \right)
  \right] .
\end{multline}
Integrating over $R_f$ gives
\begin{multline}
\int_0^\infty p_{\rm LE}(R_f|d) {\rm d}R_f =  \frac{R_b \hat{b}(x_N)}{(1-\hat{F}(x_N))} {\rm e}^{-(1-\hat{B}(x_N)) R_b}\\
%\exp\left[-(1-\hat{B}(x_N))R_b\right] \\ 
\times \left(\frac{\hat{f}(x_N)}{(1-\hat{F}(x_N)) R_b \hat{b}(x_N)} + 1 \right)
\end{multline}
and so the normalised posterior is
\begin{multline}
p_{\rm LE}(R_f|d) = \frac{(1-\hat{F}(x_N))}{1+\Lambda} \left( 1+ R_f (1-\hat{F}(x_N))\Lambda \right) \\
\times \exp\left[-R_f (1-\hat{F}(x_N))\right]
\label{loudest-eq}
\end{multline}
in which we have defined
\begin{equation}
\Lambda \equiv \frac{\hat{f}(x_N)}{(1-\hat{F}(x_N)) R_b \hat{b}(x_N)}.
\end{equation}
With the further identification $\mu \equiv R_f$ and $\hat{\epsilon} \equiv 1 - \hat{F}(x_N)$, this is Eq.~(18) of~\citep{Biswas2009} and we have shown how their parameter $\Lambda$ is related to the foreground and background distributions used here.

@@

\subsection{Foreground dominated statistic}
If we set the threshold for including an event, $x_{\rm min}$,
sufficiently high, we can ensure that $\hat{f}(x_i,\theta) \gg
\hat{b}(x_i,\theta)$ for all ranking statistics $x_i$ in the data
set.  If we can further be confident that $R_f \hat{f}(x_i,\theta) \gg R_b \hat{b}(x_i,\theta)$ for all events, then the posterior can be approximated by
\begin{multline}
p_{\rm FD}\left( R_f, R_b, \theta | d \right) \\ \propto \prod_{i} \left[
  \hat{f}\left(x_i, \theta\right)\right] R_f^N \exp\left[-\left( R_f +
  R_b \right) \right] \frac{p(\theta)}{\sqrt{R_f R_b}}.
\end{multline}
Note that these are posteriors on the number of events expected \emph{above the
threshold $x_{\rm min}$}.  The threshold choice for the foreground-dominated statistic could be different from the threshold choice applied elsewhere.  If rates $R_{f,1}$ and $R_{f,2}$ are estimated for thresholds $x_{\rm min}=x_1$, $x_{\rm min}=x_2$ respectively, these rates should be compared by equating $R_{f,1} (1-\hat{F}(x_2,\theta))$ and $R_{f,2} (1-\hat{F}(x_1,\theta))$.

Normalization over $R_b$ gives a constant factor and the posterior on
the foreground rate becomes
\begin{multline}
p_{\rm FD}\left( R_f, \theta | d \right) \\ \propto \prod_{i} \left[
  \hat{f}\left(x_i, \theta\right)\right] R_f^{N-\frac{1}{2}}
\exp\left[-R_f \right] p(\theta).
\end{multline}
Ignoring the dependence on $\theta$, this is peaked at a rate $R_f =
N-1/2$, so we have the expected result that, in the foreground
dominated regime, the rate is approximately equal to the number of
events observed (the $1/2$ comes from our use of the Jeffreys prior on
the rate).

@@

To explain this, we will use the same model as described in the
previous subsection, and we will begin with a very specific example.
Let $\Lambda_f = 1$, meaning that the expected number of actual events
with $x > x_1$ is one.  Then there is a $1/64$ chance ($1-\hat{F}(x_{LE}) \approx
1.6\%$) that the loudest event will have $x_{LE} > 4 x_1$.  Consider
this case, and let us also assume that there are no events (noise or
actual) with $x_1 < x < x_{LE}$.

The loudest event estimate basically ``throws away'' the information that there are no events in this interval.  The maximum of the loudest-event-statistic posterior on $R_f$,  Eq.~(\ref{loudest-eq}), is at $R_f = \frac{\Lambda - 1}{ \Lambda (1-\hat{F}(x_{LE}))}$.  If the value of $\Lambda$ is sufficiently high at $x_1$ (and $\Lambda$ will be even greater at $x_{LE}$), then, for this data set, we would estimate $R_f \approx \frac{1}{1-\hat{F}(x_{LE})} \gtrsim 64$.  Thus, for our assumed shape of the foreground distribution, we will estimate the rate of events above $x_1$ to be $64$ times the true rate!

Now, if the true rate really were $\Lambda_f = 64$, then the expected
number of events with $x > x_1$ would be $64$.  So in this case, the
loudest event estimate ignores the fact that there are $\sim 56 -72$
``missing'' events.  However a Bayesian estimate with $x_\mathrm{th}$
set to $x_1$ incorporates this information quite naturally, and so
(correctly) yields an estimated $\Lambda_f$ of order one.
